\textbf{ Matching forecast time-scale with ecological process }

### 2019-05-07 doc/manuscript.tex
Finished very rough draft

\title{\textbf{ Forecast time-scale determines which processes to model }}

@@

Our goal is to consider a hypothesis that, in our view, has not been carefully considered by either the push for near-term forecasting or
ongoing debates about competing modeling approaches. We expect that short and long-term forecasts will require models
incorporating different kinds of processes and fit to different kinds of data sets. We demonstrate this idea using two simulation case studies, the first for an 
eco-evolutionary scenario, the second concerned with feedbacks between community composition and ecosystem functioning.  
Our analyses show that: 
1. For short-term forecasts, phenomenological time-series approaches are hard to beat.
2. Long-term forecasts require understanding slow processes such as evolutionary and ecological selection as well as dispersal.
3. Different kinds of data reflect the operation of different processes: longitudinal data capture autocorrelation and responses of current assemblages, while 
data spanning spatial gradients capture the long-term outcome of selection and dispersal.  As the time-scale of the forecast increases, the importance of these processes shifts.
4. A key challenge for future research is determining the rate at which slow processes first influence and ultimately dominate dynamics.

@@

To demonstrate these dynamics, we simulated a diploid annual plant population at a colder than average site. During the baseline period, the population is dominated by the cold-adapted genotype. We use the simulated data from this baseline period to fit an empirical model that assumes no knowledge of the underlying eco-evolutionary process. This empirical ``temporal model" predicts population growth rate as a function of annual temperature and population size (Fig. \ref{fig:temporal_model}). We then imposed a period of warming, followed by a final period of higher, but stationary, temperature (Fig. \ref{fig:forecast} top). With the onset of warming, the population crashes as the cold-adapted genotype decreases in abundance. Eventually, frequencies of the heterozygous genotype and the warm-adapted homozygous genotype begin to increase and the population recovers (Fig. \ref{fig:forecast} bottom).

@@

The implications for forecasting are clear: In the initial stages of the warming trend, the temporal model fit to the original, baseline conditions should make better the best predictions 
about resulting population trends, but in the longer term, as evolutionary change occurs, the spatial model ultimately makes better predictions. 
To combine the spatial and temporal model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega * T(N_{t-1},K_t) + (1-\omega) * S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$ and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see Appendix for full description of the approach). The temporal model accurately predicts the impact of the initial warming trend, but eventually becomes far too pessimistic, while the spatial model does not handle the initial trend but accurately predicts the eventual, new steady state (Fig. \ref{fig:forecast} bottom). The combined model initially reflects the temporal model, but then rapidly transitions to reflect the spatial model. The rapid transition in the weighting term, $\omega$, occurs during the period of most rapid change in genotype frequencies (Fig. \ref{fig:forecast_supp}).

@@

The combined model's predictions look impressively accurate, but that is because we used the full, simulated time series to fit the weighting term. 
Unlike our spatial and temporal model forecasts, we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from from being dominated by interannual 
variation to being dominated by steady-state equilibrium. To generate a true forecast from the combined model, we would need an independent 
method to determine how the model weights shift over time.

@@

In the previous example, evolutionary change was the slow process that determined how quickly the combined model shifted from a temporal to spatial forecast. The process of compositional turnover--change in the abundances and identities of species interacting at a site--can have a similar effect at the level of communities and even ecosystems. 
We demonstrate this using a metacommunity model developed by Alexander et al. (ref). The model describes Lotka-Volterra competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient. Composition varies along the gradient because of a trade-off between growth rate and cold tolerance: cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. Sites are linked by dispersal: a specified fraction of offspring of each species leaves the site where they were produced and reaches all other sites with equal probability.

As in our eco-evolutionary example, we simulate a baseline period with variable but stationary temperature, followed by a period of rapid temperature increase and then a final period of stationary temperature. Interannual variation in temperature is the same at all sites, but mean temperature varies among sites. All sites experienced the same increase in mean temperature. We focus on two response variables at one site in the middle of the gradient. : 1) total community biomass and 2) the biomass of one focal species that dominates the baseline period

Dynamics during the baseline period reveal very strong spatial patterns across the mean temperature gradient contrasted by weak responses to interannual variation in temperature at one site. Individual species show classic ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:community-composition}). Time-averaged total community biomass increases strongly with temperature, because the fast-growing species at the warm sites produce more biomass, while the average biomass of our focal species shows a classic unimodal pattern across the temperature gradient (Fig. \ref{fig:community-baseline}). These spatial patterns are the basis for our ``spatial models" of the temperature-biomass relationship. However, at our focal site in the center of the gradient, both total biomass and the biomass of the focal species are quite insensitive to interannnual variation in temperature (Fig. \ref{fig:community-baseline}). Our ``temporal models" estimate this weak temperature effect, along with strong effects of biomass in the previous year (e.g., lags).

@@

We can forecast the effect of temperature increase on total biomass and the focal species' biomass using the spatial models and the temporal models. For total biomass, the spatial and temporal models both predict immediate increases in biomass with warming, because all species respond positively to temperature increases. However, over time, as faster growing species increase in abundance at the focal site, biomass continues to increase, approaching the predictions of the spatial model (Fig. \ref{fig:community-forecasts} middle). The predictions of the spatial and temporal model for focal species biomass contrast more dramatically, with the temporal model predicting a large increase in biomass and the spatial model predicting a decrease. Initially, the simulated abundances follow the increase predicted by the temporal model, but as faster-growing species increase in abundance, the biomass of the focal species then decreases, eventually falling below its baseline level (Fig. \ref{fig:community-forecasts} bottom). Once again, we also fit a combined model as a weighted average of the spatial and temporal model, with the weights changing over time. For total biomass (the sum across all species), the weights for the combined model shift rapidly to favor the spatial model, whereas for focal species biomass, the weights shift more slowly towards the spatial model, as the biomass of the focal species decreases and its competitors' biomass increase (Fig. \ref{fig:community-weights}).

@@

>>>>

\title{\textbf{ Matching forecast time-scale with ecological process }}

@@

Our goal is to consider a hypothesis that, in our view, has not been carefully considered by either the push for near-term forecasting or
ongoing debates about competing modeling approaches. We expect that short and long-term forecasts will require models
incorporating different kinds of processes and fit to different kinds of data sets. We demonstrate this idea using two simulation case studies, the first for an 
eco-evolutionary scenario, the second concerned with feedbacks between community composition and ecosystem functioning. Our analyses show that: 
\begin{enumerate}
	\item For short-term forecasts, phenomenological time-series approaches are hard to beat, whereas longer-term forecasts require understanding slow 
	processes such as evolutionary and ecological selection as well as dispersal.
	\item Different kinds of data reflect the operation of different processes: longitudinal data capture autocorrelation and responses of current assemblages, while 
	data spanning spatial gradients capture the long-term outcome of selection and dispersal.  Whether predictive models should be trained using longitudinal or
	spatial data sets, or both, depends on the time-scale of the desired forecast.
	\item A key challenge for future research is determining the rate at which slow processes first influence and ultimately dominate dynamics.
\end{enumerate}

@@

To demonstrate these dynamics, we simulated a diploid annual plant population at a colder than average site. During the baseline period, the population is dominated by the cold-adapted genotype. We use the simulated data from this baseline period to fit an empirical model that assumes no knowledge of the underlying eco-evolutionary process. This empirical ``temporal model" predicts population growth rate as a function of annual temperature and population size (Fig. \ref{fig:temporal_model}). We then imposed a period of warming, followed by a final period of higher, but stationary, temperature (Fig. \ref{fig:forecast} top). With the onset of warming, the population crashed as the cold-adapted genotype decreased in abundance. Eventually, frequencies of the heterozygous genotype and the warm-adapted homozygous genotype began to increase and the population recovered (Fig. \ref{fig:forecast} bottom).

@@

The implications for forecasting are clear: In the initial stages of the warming trend, the temporal model fit to the original, baseline conditions should make better the best predictions 
about resulting population trends, but in the longer term, as evolutionary change occurs, the spatial model ultimately makes better predictions. 
To combine the spatial and temporal model into a single forecast, we fit a weighting parameter, $\omega$, which varies over time and is bounded between 0 and 1. At any time point, $t$, the combined population forecast is $\omega * T(N_{t-1},K_t) + (1-\omega) * S(K_t) $ where $T$ is the temporal model, which depends on population size, $N$ and expected temperature, $K$, and $S$ is the spatial model, which depends only on $K$ (see Appendix for full description of the approach). The temporal model accurately predicted the impact of the initial warming trend, but eventually became far too pessimistic, while the spatial model did not handle the initial trend but accurately predicted the eventual, new steady state (Fig. \ref{fig:forecast} bottom). The combined model initially reflected the temporal model, but then rapidly transitioned to reflect the spatial model. The rapid transition in the weighting term, $\omega$, occured during the period of most rapid change in genotype frequencies (Fig. \ref{fig:forecast_supp}).

@@

The combined model's predictions look impressively accurate, but that is because we used the full, simulated time series to fit the weighting term. 
Unlike our spatial and temporal model forecasts, we did not generate out-of-sample predictions from the combined model; it merely provides 
a convenient way to quantify how rapidly dynamics shift from from being dominated by interannual 
variation to being dominated by steady-state equilibrium. To generate a true forecast from the combined model, we would need an independent 
method to predict how the model weights shift over time.

@@

In the previous example, evolutionary change was the slow process that determined how quickly the combined model shifted from a temporal to spatial forecast. 
The process of compositional turnover--change in the abundances and identities of species interacting at a site--can have a similar effect at the level of 
communities and even ecosystems. We demonstrate this using a metacommunity model developed by Alexander et al. (ref). The model describes Lotka-Volterra 
competitive interactions among plants within sites that are arrayed along an elevation and temperature gradient. Composition varies along the gradient because of a 
trade-off between growth rate and cold tolerance: cold sites are dominated by slow-growing species that can tolerate low temperatures, while warm sites are dominated by 
fast-growing species that are cold intolerant. Multiple species can coexist within sites because all species experience stronger competition from conspecifics than from heterospecifics. 
Sites are linked by dispersal: a specified fraction of each species' offspring leaves the site where they were produced and reaches all other sites with equal probability.

As in our eco-evolutionary example, we simulated a baseline period with variable but stationary temperature, followed by a period of rapid temperature increase, and then a 
final period of stationary temperature. Interannual variation in temperature is the same at all sites, but mean temperature varies among sites. All sites experienced the 
same absolute increase in mean temperature. We focused on two response variables at one site in the middle of the gradient: 1) total community biomass and 
2) the biomass of one focal species that dominated the baseline period.

Dynamics during the baseline period revealed very strong spatial patterns across the mean temperature gradient. 
Individual species showed classic ``Whittaker" patterns of abundances across the gradient (Fig. \ref{fig:community-composition}), while time-averaged total community 
biomass increased strongly with temperature, because the fast-growing species at the warm sites produced more biomass. 
The average biomass of our focal species showed a unimodal pattern across the temperature gradient (Fig. \ref{fig:community-baseline}). 
These spatial patterns are the basis for our ``spatial models" of the temperature-biomass relationship.

In contrast to the strong spatial patterns, population and community responses to interannual variation in temperature within sites were weak.
At our focal site in the center of the gradient, both total biomass and the biomass of the focal species were quite insensitive to interannnual variation in temperature (Fig. \ref{fig:community-baseline}). Our ``temporal models" estimate this weak temperature effect, along with strong lag effects of biomass in the previous year.

@@

We forecasted the effect of temperature increase on total biomass and the focal species' biomass using the spatial models and the temporal models. 
For total biomass, the spatial and temporal models both predicted immediate increases in biomass with warming, because all species respond positively to temperature increases. 
However, over time, as faster growing species increased in abundance at the focal site, biomass continued to increase, approaching the predictions of the spatial model 
(Fig. \ref{fig:community-forecasts} middle). The predictions of the spatial and temporal model for focal species biomass contrasted more dramatically, with the temporal 
model predicting a large increase in biomass and the spatial model predicting a decrease. Initially, the simulated abundances followed the increase predicted by the 
temporal model, but as faster-growing species increased in abundance, the biomass of the focal species then decreased, eventually falling below its baseline level 
(Fig. \ref{fig:community-forecasts} bottom).

As in the eco-evolutionary example, we also fit a combined model as a weighted average of the spatial and temporal model, with the weights changing over time. 
For total biomass (the sum across all species), the weights for the combined model shifted rapidly to favor the spatial model, whereas for focal species biomass, 
the weights shifted more slowly towards the spatial model, as the biomass of the focal species decreased and its competitors' biomass increased (Fig. \ref{fig:community-weights}). 
Again, the combined model matches the simulated data because the weights were determined by fitting directly to those data. A true, combined forecast would require
a method to determine the model weights \emph{a priori}.

@@

\subsection*{Discussion}

Our simulations demonstrate three lessons:

\emph{Lesson 1: For short-term forecasts, phenomenological time-series approaches are hard to beat, whereas longer-term forecasts require understanding slow 
processes such as evolutionary and ecological selection as well as dispersal.} [Ethan, I am hoping you can fix and add to this paragraph, perhaps
drawing from examples in other fields?] In the short-term, time-series approaches work well because lag effects are strong: the state of the system in the next 
time step depends heavily on the current state. Furthermore, in the short-term we usually are not extrapolating far into novel conditions; the historical range of variation 
may be a good guide when there has been little time for genotype frequencies or species composition to change. In contrast, farther we forecast into the future, 
the weaker the influence of the current state, and the more likely that slow processes like selection and dispersal will alter genotype frequencies and
species composition.

This lesson has implications for the emerging field of ecological forecasting. If we only focus on iterative, near-term forecasting, we are likely to find 
that time-series approaches work well, and that consideration of slow processes like evolution, dispersal, and compositional turnover rarely improves
forecast skill. Assuming that the same models will make good longer-term forecasts would be a mistake.

We do not mean to undermine the call for an emphasis on near-term forecasting. Building capacity to make near-term forecasts is extremely important, not
only because of the many critical applications (examples?) but because it is a necessary step to building a disciplinary culture and practice of ecological 
forecasting (cite Dietze?). It is also feasible, because we have the data to validate forecasts at short time-scales. Rather, the point is that near-term forecasting
alone is not enough. We also need to build capacity to make forecasts at time-scales where slow ecological processes cannot be ignored. The challenge 
will be finding ways to validate long-term predictions. Possibilities include hindcasting with paleoecological and long-term historical data sets, use of model
systems with short-generation times, and....simulations...? (or move this last bit to Lesson 3?).

\emph{Lesson 2. Different kinds of data reflect the operation of different processes; the choice of data for model training should depend on the time-scale of the desired forecast}.
Longitudinal data capture autocorrelation and responses of current assemblages, while data spanning spatial gradients capture the long-term outcome of 
selection and dispersal. For short-term forecasts, building models using longitudinal data is clearly the best choice. As our simulations show, models fit to 
steady-state, spatial patterns make terrible predictions in the short term. Conversely, models based on longitudinal data make poor predictions in the long-term,
while the models fit to spatial patterns can perform well. Although the temporal and spatial models we fit are phenomenological, they capture different kinds
 of underlying processes. The traditional debate about phenomenological vs. process-based models obscures the fact that, depending on the choice of the
 training data, phenomenological models can represent different kinds of processes.

 Why not simply build a process-based model? Perhaps something like the process-based models we used to generate the simulated time series in the first place? 
 That kind of model could make good predictions at both short and long time-scales. Unfortunately, this approach is just not feasible in most cases.
 The first challenge is simply determining the critical process. Take our eco-evolutionary example: should we simply assume that every annual plant species
 has warm and cold-adapted genotypes? And that temperature adaptation will be the key trait driving the response to warming, as opposed to the effects of 
 compositional turnover highlighted in our second example? Second, even if the key process is known in advance, fitting a mechanistic model presents its
 own set of challenges. The model for our eco-evolutionary example requires, in order of increasing empirical difficulty, demographic rates for the full life
 cycle, including density dependence, a quantitative mapping of genotype-by-environment interactions onto fitness, and dispersal kernels. The model for 
 our community and ecosystem example requires demography, environment-dependent intra- and interspecific competition coefficients, and dispersal kernels.
 We can think of empirical models that capture some of these features, but not all of them. Finally, even were it possible to fit such a model, the high number of 
 parameters and associated parameter error might lead to poor predictive skill (refs?).

Inevitably, models used for ecological forecasting will be partially if not entirely phenomenological. But that does not mean that they will not rely on process-based
understanding in some way. The point of our simulations is that different processes should be considered for different forecast time-scales, and this can be done 
by fitting phenomenological models to different kinds of data sets. Even when process-level understanding does not enable a fully mechanistic model, it can
improve the specification of phenomenological models.

\emph{Lesson 3. A key challenge for future research is determining the rate at which slow processes first influence and ultimately dominate dynamics.}
Our simulated examples suggest the potential for a combined forecast, which weights predictions from spatial and temporal models, to 
accurately forecast both short and long-term dynamics. But we cheated: our combined ``forecasts'' were fit to data, not generated prospectively
using \emph{a priori} model weights. Finding a general solution for determining those weights poses a fascinating research problem whose 
solution would lead to important applications in ecological forecasting.

We see a role for new theory and new empirical tests. On the theoretical side, what are the parameters that determine how quickly slow processes
influence dynamics? We can play with our simulation models to provide some qualitative answers: the stronger the selection (at the level of genotype 
or species), the shorter the lifespan of the focal organisms, and the greater the dispersal, the more rapid the shift from the temporal to the spatial model
(should we actually DO these runs?). But providing general, analytical solutions would be difficult, even for our relatively simple models.

On the empirical side, the key problem is finding or creating data sets to estimate the weighting function. As described under Lesson 1,
this requires data sets appropriate for validating long-term forecasts.
(Move material from end of  Lesson 1 down here?). With enough case studies, we might be able to infer patterns
that could guide applications in new systems.

Anything else to add?
